Xolotl: An Intuitive and Comprehensible Neuronal Simulator

### 2018-05-22 PaperLatex/frontiers.tex
Add descriptions of some figures

%%% Leave the Abstract empty if your article does not require one, please see the Summary Table for full details.
\texttt{xolotl} is an open-source neuronal simulator written in \texttt{C++} with \texttt{MATLAB} wrappers. Complex models and networks can be designed efficiently using an intuitive language tightly coupled to the object-based architecture of the underlying \texttt{C++} code. Models are specified by adding conductances to compartment objects. The structure is modular, serialized, and searchable, permitting high-level programmatic control over nearly all features of the models. \texttt{C++} templates are provided for developing new conductances, compartments, and integration schemata. \texttt{xolotl} readily implements parallel-processing on multicore processors and high-performance computing clusters. It also includes a customizable graphical user interface (`puppeteer`) for rapid prototyping and hand-tuning conductances in real-time. In addition, \texttt{xolotl} comes packaged with a powerful optimization toolbox (\texttt{procrustes}) for optimizing any model parameters accessible in the \texttt{xolotl} tree. The modular structure and accessibility to all parameters, variables, and dynamics of the model network in \texttt{MATLAB} facilitate interoperability with other specifications (viz. \texttt{NeuroML, SBML}), simulators (viz \texttt{NEURON, Brian, NEST}), and web-based applications (viz. \texttt{Geppetto}). \texttt{xolotl} is freely available at \url{https://github.com/sg-s/xolotl}. This tool provides rapid implementation and fast simulation of neuronal models while permitting full control over every aspect of the network and integration.

@@

For Original Research Articles \citep{conference}, Clinical Trial Articles \citep{article}, and Technology Reports \citep{patent}, the introduction should be succinct, with no subheadings \citep{book}. For Case Reports the Introduction should include symptoms at presentation \citep{chapter}, physical exams and lab results \citep{dataset}.

@@

@@

@@

\FloatBarrier

\section{Nomenclature}

\subsection{Resource Identification Initiative}
To take part in the Resource Identification Initiative, please use the corresponding catalog number and RRID in your current manuscript. For more information about the project and for steps on how to search for an RRID, please click \href{http://www.frontiersin.org/files/pdf/letter_to_author.pdf}{here}.

\subsection{Life Science Identifiers}
Life Science Identifiers (LSIDs) for ZOOBANK registered names or nomenclatural acts should be listed in the manuscript before the keywords. For more information on LSIDs please see \href{http://www.frontiersin.org/about/AuthorGuidelines#InclusionofZoologicalNomenclature}{Inclusion of Zoological Nomenclature} section of the guidelines.

\section{Additional Requirements}

For additional requirements for specific article types and further information please refer to \href{http://www.frontiersin.org/about/AuthorGuidelines#AdditionalRequirements}{Author Guidelines}.

The Author Contributions section is mandatory for all articles, including articles by sole authors. If an appropriate statement is not provided on submission, a standard one will be inserted during the production process. The Author Contributions statement must describe the contributions of individual authors referred to by their initials and, in doing so, all authors agree to be accountable for the content of the work. Please see  \href{http://home.frontiersin.org/about/author-guidelines#AuthorandContributors}{here} for full authorship criteria.

\section*{Acknowledgments}
This is a short text to acknowledge the contributions of specific colleagues, institutions, or agencies that aided the efforts of the authors.

\section*{Data Availability Statement}
The datasets [GENERATED/ANALYZED] for this study can be found in the [NAME OF REPOSITORY] [LINK].
% Please see the availability of data guidelines for more information, at https://www.frontiersin.org/about/author-guidelines#AvailabilityofData

\begin{figure}[h!]
\begin{center}
\includegraphics[width=10cm]{logo1}% This is a *.eps file
\end{center}
\caption{ Enter the caption for your figure here.  Repeat as  necessary for each of your figures}\label{fig:1}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics[width=15cm]{logos}
\end{center}
\caption{This is a figure with sub figures, \textbf{(A)} is one logo, \textbf{(B)} is a different logo.}\label{fig:2}
\end{figure}

%%% If you are submitting a figure with subfigures please combine these into one image file with part labels integrated.
%%% If you don't add the figures in the LaTeX files, please upload them when submitting the article.
%%% Frontiers will add the figures at the end of the provisional pdf automatically
%%% The use of LaTeX coding to draw Diagrams/Figures/Structures should be avoided. They should be external callouts including graphics.

>>>>

%%% Leave the Abstract empty if your article does not require one, please see the Summary Table for full details.
\texttt{xolotl} is an open-source neuronal simulator written in \texttt{C++} with \texttt{MATLAB} wrappers. Complex models and networks can be designed efficiently using an intuitive language tightly coupled to the object-based architecture of the underlying \texttt{C++} code. Models are specified by adding conductances to compartment objects. The structure is modular, serialized, and searchable, permitting high-level programmatic control over nearly all features of the models. \texttt{C++} templates are provided for developing new conductances, compartments, and integration schemata. \texttt{xolotl} readily implements parallel-processing on multicore processors and high-performance computing clusters. It also includes a customizable graphical user interface (`puppeteer`) for rapid prototyping and hand-tuning conductances in real-time. In addition, \texttt{xolotl} comes packaged with a powerful optimization toolbox (\texttt{procrustes}) for optimizing any model parameters accessible in the \texttt{xolotl} tree. The modular structure and accessibility to all parameters, variables, and dynamics of the model network in \texttt{MATLAB} facilitate rapid construction and assessment of model networks. \texttt{xolotl} is freely available at \url{https://github.com/sg-s/xolotl}. This tool provides straightforward implementation and fast simulation of neuronal models while permitting full control over every aspect of the network and integration.

@@

\texttt{xolotl} (\url{https://github.com/sg-s/xolotl}) is a fast single-compartment and multi-compartment simulator in \texttt{C++} with \texttt{MATLAB} wrappers (\url{https://www.mathworks.com/products/matlab.html}). Written with an emphasis on flexibility and speed, \texttt{xolotl} can simulate single-compartment conductance-based models, networks of these, and detailed multi-compartment models. \texttt{xolotl} exploits a novel automatic type system, \texttt{cpplab}, which binds \texttt{MATLAB} code to \texttt{C++} header files, creating objects and classes \textit{ad libitum} in \texttt{MATLAB} which reflect the underlying object-oriented code. \texttt{xolotl} implements \texttt{cpplab} to represent the nested structure of conductance-based models, and exploits the computational efficiency of the low-level programming language to quickly integrate models. For this reason, models can be implemented entirely in \texttt{MATLAB} with few lines of code.

Models are specified in \texttt{MATLAB} by a \texttt{xolotl} object which contains compartment objects which themselves contain conductances. Synapses belong to the \texttt{xolotl} object and connect compartments together. The high-level specification supports arbitrarily large network and multi-compartment morphologies.

\texttt{xolotl} provides parameter optimization capabilities through the algorithm-agnostic \texttt{procrustes} toolbox. Any network parameters accessible through the \texttt{xolotl} structure can be optimized using arbitrary algorithms and objective functions on multi-core computers and high-performance computing clusters.

The software has been implemented in \texttt{MATLAB} due to its ease-of-use and popularity among neuroscientists. \texttt{cpplab} provides a powerful backend for specifying and integrating models without relying on the significantly slower and limiting \texttt{MATLAB codegen}. Minimal experience with \texttt{MATLAB} is required to use \texttt{xolotl}, and all equations and integration methods are provided transparently to the end user. No string parsing of equations is required.

\texttt{xolotl} comes packaged with visualization functions, a graphical user interface for real-time manipulation of model parameters. Plotting of voltage, intracellular calcium, conductance gating functions, and time constants is provided by built-in \texttt{xolotl} methods. The \texttt{puppeteer} toolbox permits real-time tuning of any network parameters using numerical sliders in a graphical interface which displays the resultant membrane potential and intracellular calcium traces. The ease-of-use of these tools lends them to pedagogical applications and rapid exploration of toy models. This tool aims to simplify the investigation of dynamics of complex neural network models, facilitate collaborative modeling, and complement other tools being developed in the neuroinformatics community.

@@

\section{Worked Examples}

Using \texttt{xolotl} in \texttt{MATLAB}, users create a \texttt{xolotl} object and populate it with compartments, synapses, and controllers. Each field is a \texttt{cpplab} object constructed by a function call to \texttt{add}. The model is integrated with the \texttt{integrate} function where the membrane potential, intracellular calcium concentration, controller states, intrinsic currents, and synaptic currents can be outputs.

\texttt{xolotl} comes packaged with a library of pre-existing conductance and synapse objects which greatly simplify the task of constructing model neurons. These objects can be referenced by name and added directly to a compartment. Novel conductance dynamics can be easily written by modifying a template header file contained in the \texttt{xolotl} distribution.

\subsection{Simulating a Hodgkin-Huxley Model}

The seminal Hodgkin-Huxley model contains a fast inactivating sodium conductance which promotes spiking, a non-inactivating potassium delayed rectifer, and a passive leak current (\ref{fig:figurehh}A). A compartment named \texttt{HH} with compartment properties of membrane capacitance $C_m = 10 ~ \mu\mathrm{F/mm^2}$ and surface area $A = 0.1 ~ \mathrm{mm^2}$ can be specified by \ref{fig:figurehh}B. Compartment, conductance, synapse, and controller properties can be specified during the call to the \texttt{add} function, or after construction, using dot-notation in \texttt{MATLAB} (e.g. \texttt{x.HH.Cm}). \ref{fig:figurehh}C shows the \texttt{MATLAB} command prompt after invoking the \texttt{xolotl} object \texttt{x}, displaying the hierarchical structure inherent in conductance-based treatments of neurodynamics.

This model was constructed using conductances from Liu \textit{et al} 1998. In the absence of applied positive current, the model is quiescent and tonically spikes under 0.2 nA of applied current (\ref{fig:figurehh}D). The \texttt{integrate} function takes the applied current as an argument (e.g. \texttt{x.integrate(AppliedCurrent)}), so that the \texttt{xolotl} object is agnostic to integration-specific perturbations. The \texttt{plot} function generates voltage and intracellular calcium traces, where the voltage trace is colored by the dominant current. If the membrane potential is increasing, the strongest instantaneous inward current colors the trace. Conversely, if the membrane potential is decreasing, the strongest outward current colors the trace instead. \ref{fig:figurehh}F-I display the results of the \texttt{show} function. Activation and inactivation steady-states and the voltage-dependent time constants of these gating variables describe the conductance dynamics in absence of other channel types.

\subsection{Performing a Voltage Clamp Experiment \textit{in-silico}}

\texttt{xolotl} can recapitulate the results of voltage clamp experiments. \ref{fig:figureclamp} displays the procedure to clamp the membrane potential of a cell with a delayed rectifier potassium conductance. The second argument of the \texttt{integrate} function determines the clamped voltage (e.g. \texttt{x.integrate([], VoltageClamp)}). Isolated currents under voltage clamp approach the steady-state (\ref{fig:figureclamp}D-E) so that a current-voltage relation at steady-state can be extracted (\ref{fig:figureclamp}F). The derivative of the IV curve is the steady-state conductance (\ref{fig:figureclamp}G). Fitting a sigmoid to various powers yields a model for the current dynamics (\ref{fig:figureclamp}H-I). These figures describe graphically the theoretical underpinnings of current analysis through voltage clamp and can serve as an effective pedagogical tool for computational and quantitative neuroscience.

\subsection{Simulating Network Models}

Network models in \texttt{xolotl} consist of compartment objects connected by synapses. Synapses are stored in a vector array as a field of the \texttt{xolotl} object in \texttt{MATLAB}. Presynaptic and postsynaptic labels indicate the connectivity of the synapse. Figure \ref{fig:figurenetwork} implements a model of the triphasic pyloric rhythm in the stomatogastric ganglion of crustaceans. The pyloric model contains three compartments and seven synapses (\ref{fig:figurenetwork}A). This structure is reciprocated in the hierarchy of the \texttt{xolotl} object, where conductances are contained within compartments (\ref{fig:figurenetwork}B). Synapses are upper-level properties of the network which point between two compartments (\ref{fig:figurenetwork}C). This exploits vectorized operations in \texttt{MATLAB} and does not require each synapse to possess a unique name. The \texttt{plot} function generates multiple subplots when called for a network with multiple compartments (\ref{fig:figurenetwork}D-F).

@@

AH and SG-S wrote the manuscript and the code. EM provided funding and moral support.

AH received funding from National Institute on Drug Abuse (NIDA) through the undergraduate training grant in computational neuroscience (1R90DA033463-01).

\section*{Acknowledgments}
The authors would like to thank Mara CP Rue and Hillary Rodgers for beta-testing the \texttt{xolotl} software.

\section*{Data Availability Statement}

The code to generate all figures can be found in the \texttt{xolotl} repository (\url{https://github.com/sg-s/xolotl}).

@@
